Denotational Semantics of General Payment Primitives, and Its Payment System

### 2022-10-04 packages/spec-haskell/yellowpaper/Paper1.tex
fixed a round of proof-readings

\begin{abstract}
Payment systems in information age are still largely modeled after their analog predecessors. While
some electronic money payment systems do utilize computing technology and Internet, this paper
presents a case that a true modernization can still be done by (a) making payments happening
continuously over time (b) involving more than two parties in a payment if necessary (c) having
compositional financial contracts.

This paper first explores the foundation of modern payment system, which consists of a money
distribution model, payment primitives, payment execution systems of financial contracts and
different forms of money mediums. Then the paper uses \textit{denotational semantics} to formally
define payment primitives for modern payment system. Lastly, this paper also includes an overview
of \textit{Superfluid protocol}, a reference implementation of the payment primitives and a its
payment system.

This paper also includes literate Haskell code that compiles as the constructive artifact for the
formal specifications.

@@

It should be fair to say, every aspect of money are controversial: the nature of money, the value of
money, money and banking, and monetary reconstruction. Two major schools of thoughts about theory of
money are the \textit{Austrian school} (\cite{von2013theory}) and the \textit{Chicago school}
(\cite{friedman1989quantity}). That is before the appearance of Internet-era version of monetary
reconstruction, broadly defined as cryptocurrency, which challenges theories of money further and
demands their updates (\cite{ammous2018can} \cite{hardle2020understanding}).

This yellow paper does not intend to address these controversies, but to focus on the function of
money. According to Von Mises:

@@

This paper adds a new controversy to money, that is to present a survey challenging the preconceived
notion of how money can perform its function of medium of exchange.

In part\ref{part:foundation}, we shall first explore the foundation for the survey. Here we present
a formal definition of payment system and its components. We then select a few relevant approaches
used in computer science useful for modeling and defining formal specification for the payment
system.

One of the approaches is \textit{denotational semantics}, which is used in part \ref{part:gpp} of
the paper to define the general payment primitives. Along with the denotational semantics, a
restatement\footnote{It is a borrowed term from common law: ``restatement of the law''. In our case,
the denotative mathematical laws.} of it in \textit{Haskell programming language}
(\cite{hudak1992report} \cite{jones2003haskell} \cite{marlow2010haskell}) is also included.

In part \ref{part:sf}, a reference implementation of the general payment primitives and its payment
system called \textit{Superfluid Protocol} is introduced.

@@

It can be solely defined by these components:

@@

    \item \textit{payment execution environment} performs payment primitives,

@@

\begin{itemize}
    \item Money can be distributed continuously over time, as opposed to be discrete.

@@

This discrete nature of this money distribution model is schematically depicted
in figure \ref{fig:discrete-md}:

@@

\subsubsection{Using Context}

@@

Before first time in this paper listing specification in Haskell, here are some highlights of
conventions in the style of the code used.

@@

@@

In order to make the specification free of specific choices of core data types in
implementation, \textit{FunctionalDependencies}, \textit{TypeFamilies}
and \textit{TypeFamilyDependencies} are extensively used. And in order to make the code less
cluttered, type quality operator $(\sim)$
used. Here is an example:

@@

A payment primitive is a data type with a generator that creates payment primitive data from the
shared context $ctx$, primitive specific argument $args$ and timestamp, along with an update of the
shared context. Here is an example of generator for payment primitive \textit{a}:

Payment primitive data then can be used to create a delta update of money
distribution\footnote{Specifically being monoidal, that is in short a set that has associative
binary operation and an identity element. See https://ncatlab.org/nlab/show/monoid.}

@@

The best known primitive is instant transfer of monetary value between one monetary unit to
another. The introduction of context enables more primitives to be defined, and this will be
discussed in later chapters.

@@

It is out of scope for this paper to survey in-depth the problem space of the operation semantics of
payment execution environments. Nonetheless, a simplified model and some potential extensions to it
is discussed briefly.

@@

While money distribution does not assign meaning to the range of monetary values, in real world
applications, negative values can have special meanings. In the following analysis, we call any
money unit that has a negative monetary value insolvent.

@@

To mitigate this uncertainty, a concept called buffer is introduced. A buffer has a monetary value
which is set aside in a solvency conditional financial contract, such that if a solvency condition
arises, the buffer maybe drawn to cover the loss introduced by the nondeterministic timing of the
execution.

@@

But it introduces a different type of systemic risk, that is denial-of-service, since the system
cannot advance the system time until all next executable contracts are executed.

@@

The focus of this paper is to formally define a set of payment primitives that modernize our payment
systems. To prevent reinventing wheels, it is relevant to discuss first some approaches in computer
science that are believed to help solving this challenge.

@@

Recall that we want our modernized payment systems to handle money distribution continuously over
time, and its financial contracts compositional. A very closely related software design paradigm
best known to address those needs is \textit{functional reactive programming (FRP)}.  It was first
introduced by Conal Elliott \& Paul Hudak in solving multimedia
animations \cite{elliott1997functional}. Later on, Hudak also worked on \cite{hudak2002arrows} and
\cite{wan2000functional} making FRP a more general framework for programming hybrid system with
continuous behaviors in a high-level, declarative manner.

After FRP got more adoption, it evolved to some variations that support discrete semantics, and some
variations better suited more for interactive systems. In this paper we will stick to and revisit
the basic constructions of the original formulation used in \cite{elliott1997functional} where we
will draw inspirations from.

@@

Values that vary over continuous time are called \textit{behaviors}. They are first-class values,
and are built up compositionally. In modernized payment systems, we should want the monetary values
of monetary units capable of varying over continuous time. The semantic function of
$\alpha$-behaviors produces the value of type $\alpha$ of a behavior at a given time:

@@

In modernized payment systems, the payment primitives executed in payment execution environments are
one type of events. More types of events can be read from the original paper \footfullcite[][section
2.3 Semantics of Events]{elliott1997functional}.

@@

In the context of modernized payment systems, the occurrences of payment primitives (events) change
the behaviors of monetary units in how much monetary values it has over continuous. The building
blocks of the financial contracts should be about modeling these events and their reactivity
declaratively.

@@

The syntactic category we are interested in is \textit{payment primitives}, which we should treat
them as FRP-style \textit{Behavior} data types. In the following chapters, it is also unambiguously
referred to in short as $[\![.]\!]$. Various $[\![.]\!]$ must be compositional, \ie must be defined
by structural recursion.

It is important to note that our purpose of using the denotational semantics is to give precise
meaning to payment primitives, independent of its implementation (which deals with performance,
optimization, side effects, etc.). We will spend the next chapter exploring the denotational
semantics for general payment primitives in modernized payment systems.

@@

The specifications will be formally defined using denotational semantics, along with a restatement
in Haskell also applied as a constructive artifact friendly to computer environment\footnote{One may
argue for a restatement in \textit{Agda} instead, for it has a richer dependently-type system for
desirable constructive proofs. This matter will be dealt in the ``future investigations'' chapter of
this paper.}.

@@

    \item Semantic function $[\![.]\!] :: \mathcal{M} \rightarrow [\![\mathcal{M}]\!] $ evaluates
the syntax of money distribution, payment primitives, etc.
\end{itemize}

@@

And it must satisfy the following law:

@@

\subsection{Proportional Distribution Primitives}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

An even more powerful abstraction is to model participants involved a payment primitive
a \textit{network} (we use symbol $w$ for them).

@@

The answer is most likely a no. Not because it is not useful, after all mathematical formula is
about truth and perhaps also elegance in it, but because of we may miss optimization opportunities
needed in implementation when more specialized versions are used instead\footnote{We will not though
discuss the subjective aspects, including in software engineering principles such as YAGNI
(\url{https://en.wikipedia.org/wiki/You_aren't_gonna_need_it}), or user experience perspective of
the payment primitives.}.

In the next chapter, we will look into a reference implementation where these optimizations are
applied.

@@

Superfluid protocol (``the protocol'')\footnote{The code base is available
at \url{https://github.com/superfluid-finance/protocol-monorepo/}.} is the first implementation of
the \textit{denotational semantics of payment primitives} (though before it was formalized and named
so) on \textit{Ethereum Virtual Machine} \cite{wood2014ethereum}. The first version of the protocol
is written in Solidity programming language \footnote{\url{https://soliditylang.org/}}.

To better serve as a reference implementation of the \textit{modern payment system} formalized by
this paper and its sequels, an implementation in Haskell was created. It aims to implement the full
specifications of (a) denotational semantics of payment primitives (b) compositional financial
contracts for the payment primitives (c) token \& note model of money mediums, and their execution
environment.

This paper covers the overview of the implementation of (a).

@@

For the purpose of compositional financial contracts, it is useful to separate monetary values that
bearers can use immediately (called \textit{untappedValue} in code) from ones that are set aside for
other financial purpose. The concept of \textit{real time balance} is thus created. \textit{Real
time balance} is a functorful\footnote{Bartosz Milewski has an excellent series on these
topics: \url{https://bartoszmilewski.com/2015/01/20/functors/}, and he uses the term ``functorful''
to convey the idea of generalized container.} of monetary values. It can be converted to a single
monetary value.

Although since this is not the subject of this paper, it suffices to just show its definition
instead:

@@

\begin{itemize}
    \item preserve the program correctness (and if possible, with proof of equivalence),
    \item optimize for efficient computation,
    \item and provide a software interface for using it.
\end{itemize}

@@

\paragraph{Monetary Unit Data (MUD)}

Recall that all monetary units have a monetary value function: $$ \nu :: u \rightarrow \mathbb{N}$$.

@@

\begin{code}
class ( SuperfluidCoreTypes sft
      ) => MonetaryUnitDataClass mud sft | mud -> sft where
    -- | π function - balance provided (hear: π) by the monetary unit data.
    balanceProvided
        :: forall t rtb.
           -- indexed type aliases
           ( t ~ SFT_TS sft
           , rtb ~ SFT_RTB sft
           )
        => mud -> t -> rtb
\end{code}

@@

\paragraph{Agreement Contract}

@@

\begin{code}
-- | Agreement contract type class.
class ( SuperfluidCoreTypes sft
      , Default ac
      , MonetaryUnitDataClass ac sft
      , Default (AgreementOperationOutput ac)
      , Traversable (AgreementOperationOutputF ac) -- <= Foldable Functor
      ) => AgreementContract ac sft | ac -> sft where

    -- | ω function - apply agreement operation ~ao~ (hear: ω) to the agreement
    --                operation data ~ac~ to get a tuple of:
    --
    --   1. An updated ~ac'~.
    --
    --   2. A functorful delta of agreement monetary unit data ~mudsΔ~, which
    --      then can be appended to existing ~mudΔ~.  This is what can make an
    --      agreement scalable.
    applyAgreementOperation
        :: forall t ao aoo.
           -- indexed type aliases
           ( t ~ SFT_TS sft
           , ao ~ AgreementOperation ac
           , aoo ~ AgreementOperationOutput ac
           )
        => ac -> ao -> t -> (ac, aoo)
    data family AgreementOperation ac :: Type
    data family AgreementOperationOutputF ac :: Type -> Type
    type family AgreementOperationOutput ac = (smuds :: Type) | smuds -> ac
\end{code}

@@

Figure \ref{fig:ac-omega} is a visualization of ``machinery'' of the $\omega$ function:

\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth]{../assets/agreement-contract-omega-function.png}
    \caption{Agreement Contract $\omega$ Function}
    \label{fig:ac-omega}
\end{figure}

@@

It is also important to note that $AgreementContract$ itself is also a $MonetaryUnitDataClass$, but
it is not a semigroup, hence it is used to replace the previous agreement contract data. This has
optimization implications, as in that if agreement contracts do not produce zero balance then they
must be included in the $\nu$ function, that can make the $\nu$ function $O(N)$ to the number of
agreement contracts a monetary unit is associated with.

@@

\subsection{Lens Data Accessors}

@@

The pro-functor version of might seem very obscure at first, but its data representation is rather
self-explanatory: a lens is simply a pair of getter ($view$) and setter ($update$) encoded in the
data.

@@

\paragraph{Instant Value MUD}

@@

Note that (a) $^.$ is the view function (getter) for Lens types. (b) semigroup operator $(<>)$ defines
how two monetary unit data can be combined into one.

\paragraph{Constant Flow Agreement (CFA) MUD}

@@

\paragraph{Decaying Flow Agreement (CFA) MUD}

@@

\paragraph{Universal Index (UIDX)}

@@

\paragraph{Proportional Distribution Index (PDIDX)}

@@

\subsection{Hierarchy of Agreements}

To finish, figure \ref{fig:ac-agreement-data-structures} is the illustration of data structures for
the denotational payment primitives implemented with agreement framework:

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{../assets/agreement-data-structures.png}
    \caption{Agreement Data Structures}
    \label{fig:ac-agreement-data-structures}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\part*{Notes on Future Investigations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

One of the advantages of using Haskell language was chosen to write the reference implementation is
that it is a language with industrial strength, and the implementation could be easily integrated
into a production code base without much performance compromises.

However, a major disadvantage is that it is not equipped with sufficient apparatus for program
correctness proofs. The best it can offer without engaging more experimental Haskell features is to
use \textit{QuickCheck}\footnote{\url{https://wiki.haskell.org/QuickCheck} - Haskell wiki page for
QuickCheck.} to the necessary laws are not violated using randomized test vector.

@@

It is based on the insight of the deep connection (equivalence/isomorphism) between logic and
dependently typed programming, often called "the Curry–Howard correspondence", as discovered and
developed during the 20th century (explained in \cite by Phillip Wadler as ``Propositions as
Types''). Agda being a dependently typed functional programming language fully embodies this
insight, hence a good candidate for creating provable correct program.

To learn more about Agda, see refer to these footnotes\footnote{\url{https://plfa.github.io/} -
Programming Language Foundations in
Agda.} \footnote{\url{https://wiki.portal.chalmers.se/agda/pmwiki.php} - Agda
Wiki.} \footnote{\url{https://wiki.portal.chalmers.se/agda/Main/Community} - Agda community page.}.

@@

The execution environments for these financial contracts are also fertile ground for new ideas, from
deterministic execution to avoid the usage of buffer, to using distributed ledger technology to
deep-embed these financial contracts in the consensus layer.

@@

The \textit{real time balance} introduced in this paper can be seen as a instance in
the \textit{general accounting domain} specifically for financial contracts in dealing with payment
primitives, a real time version of the cash flow accounting. It is a long with other types,
including balance sheet accounting, income statement accounting. The conversion between these
instances of general accounting domains are often called ``reconciliations''.

@@

@@

This paper defines the what constitutes a modern payment system formally specified using
denotational semantics and restated using Haskell programming language.

>>>>

\begin{abstract}
Payment systems in information age are still largely modeled after their analog predecessors. While
electronic money payment systems do utilize computing technology and Internet, this paper presents a
case that a true modernization can be reached by (a) making payments happening continuously over
time, (b) involving more than two parties in a payment if necessary, (c) having compositional
financial contracts.

This paper first explores the foundation of modern payment system, which consists of a money
distribution model, payment primitives, payment execution systems of financial contracts and
different forms of money mediums. Then the paper uses \textit{denotational semantics} to formally
define payment primitives for modern payment system. Lastly, this paper also includes an overview
of \textit{Superfluid protocol}, a reference implementation of the payment primitives and its
payment system.

This paper includes literate Haskell code that compiles as the constructive artifact for the formal
specifications.

This paper is the first one of the series yellow papers about modern payment system dubbed
``semantic money''.

@@

It should be fair to say, every aspect of money is controversial: the nature of money, the value of
money, money and banking, and monetary reconstruction. Two major schools of thoughts about theory of
money are the \textit{Austrian school} (\cite{von2013theory}) and the \textit{Chicago school}
(\cite{friedman1989quantity}). That is before the appearance of Internet-era version of monetary
reconstruction, broadly defined as cryptocurrency, which challenges theories of money further and
demands their updates (\cite{ammous2018can} \cite{hardle2020understanding}).

This yellow paper does not intend to address these controversies, instead it focuses on the function
of money. According to Von Mises:

@@

This paper adds a new controversy to money, that is to present a foundation of what constitutes of
a \textit{modern payment system} and a set of \textit{payment primitives} for the system, in order
to challenge the preconceived notions of how money can perform its function of medium of exchange.

In part\ref{part:foundation}, we shall first explore the foundation. Here we present a formal
definition of payment system and its components. We then select a few relevant approaches used in
computer science useful for modeling and defining formal specification for the payment system.

One of the approaches is \textit{denotational semantics}, which is used in part \ref{part:gpp} of
the paper to define the \textit{general payment primitives}. Along with the denotational semantics,
a restatement\footnote{It is a borrowed term from common law: ``restatement of the law''. In our
case, the denotative mathematical laws.} of it in \textit{Haskell programming language}
(\cite{hudak1992report} \cite{jones2003haskell} \cite{marlow2010haskell}) is also included.

In part \ref{part:sf}, a reference implementation of the general payment primitives and its payment
system called \textit{Superfluid Protocol} is introduced with an overview of it.

@@

It is solely defined by these components:

@@

    \item \textit{payment execution environment} performs payment primitives encoded
    in \textit{financial contracts},

@@

\begin{itemize}
    \item Money can be distributed continuously over time, as opposed to be in discrete chunks.

@@

This discrete nature of this money distribution model is schematically depicted
in figure \ref{fig:discrete-md}.

@@

\subsubsection{Adding Context $\gamma$}

@@

Before the first time in this paper listing specification in Haskell, here are some highlights of
conventions in the style of the code used.

\paragraph{Language Extensions}

@@

Other notable extensions used in the paper are: $FunctionalDependencies$, $TypeFamilies$,
$TypeFamilyDependencies$, $GADTs$, $MultiWayIf$.

@@

In order to make the specification free of specific choices of core data types in
implementation, \textit{FunctionalDependencies}, \textit{TypeFamilies}
and \textit{TypeFamilyDependencies} are extensively used. As a consequence, the type signature can
look very cluttered. In order to address that, type quality operator $(\sim)$ is also used. Here is
an example snippets:

@@

A payment primitive is a data type with a generator function that produces payment primitive from
the shared context $ctx$, primitive specific argument $args$ and timestamp $t$, along with an update
of the shared context. Type signature \ref{eq:genPrim} is for the generator of payment
primitive \textit{a}:

\begin{equation}\label{eq:genPrim}
    genPrim_a :: ctx \rightarrow args_a \rightarrow t \rightarrow prim \times ctx
\end{equation}

Payment primitive data then can be used to create a delta update of money
distribution\footnote{Specifically being monoidal, that is in short a set that has associative
binary operation and an identity element. See https://ncatlab.org/nlab/show/monoid.}:

@@

The most known primitive is instant transfer of monetary value between one monetary unit to
another. The introduction of context enables more primitives to be defined, and this will be
discussed in part \ref{part:gpp}.

@@

It is out of scope for this paper to survey in-depth the problem space of the operational semantics
of payment execution environments. Nonetheless, a simplified model and some potential extensions to
it is discussed briefly to place payment primitives in the big picture.

@@

While money distribution does not assign meaning to the range of monetary values, in real world
applications, negative values can have special meanings. In the following analysis, we call any
money unit that has a negative monetary value \textit{insolvent}.

@@

To mitigate this uncertainty, a concept called \textit{buffer} is introduced. A buffer has a
monetary value which is set aside in a solvency conditional financial contract, such that if a
solvency condition arises, the buffer maybe drawn to cover the loss introduced by the
nondeterministic timing of the execution.

@@

On the other hand, it introduces a different type of systemic risk, that is a kind of
denial-of-service. It is due to that the complexity of $fcUpdate$ is $O(log(n))$, the system may not
be able to advance its system time until all next executable contracts are executed.

@@

The focus of this paper is to formally define a set of payment primitives that modernize our payment
systems. To prevent reinventing wheels, it is relevant to discuss first some approaches in computer
science that are believed to help tackling this challenge.

@@

Recall that we want our modern payment system to handle money distribution continuously over time,
and its financial contracts compositional. A very closely related software design paradigm best
known to address those needs is \textit{functional reactive programming (FRP)}.  It was first
introduced by Conal Elliott \& Paul Hudak in solving multimedia animations
(\cite{elliott1997functional}). Later on, Hudak also worked on \cite{hudak2002arrows} and
\cite{wan2000functional} making FRP a more general framework for programming hybrid system with
continuous behaviors in a high-level, declarative manner.

After FRP got more adoption, it evolved to some variations that support discrete semantics, and some
variations better suited for interactive systems. In this paper we will stick to and revisit the
basic constructs of the original formulation used in \cite{elliott1997functional} where we will draw
inspirations from.

@@

Values that vary over continuous time are called \textit{behaviors}. They are first-class values,
and are built up compositionally. That is what we want in modern payment system also. The semantic
function of $\alpha$-behaviors produces the value of type $\alpha$ of a behavior at a given time:

@@

In modern payment system, the payment primitives executed in payment execution environments are
one type of events. More types of events can be read from the original paper \footfullcite[][section
2.3 Semantics of Events]{elliott1997functional}.

@@

In the context of modern payment system, the occurrences of payment primitives (events) change the
behaviors of monetary units in how much monetary values it has over continuous. The building blocks
of the financial contracts should be about modeling these events and their reactivity declaratively.

@@

The syntactic category we are interested in is \textit{payment primitives}, which we should treat
them as FRP-style \textit{Behavior} data types. In the following chapters, it is also unambiguously
referred to in short as $[\![.]\!]$. Various $[\![.]\!]$ must be compositional, \ie\ must be defined
by structural recursion.

It is important to note that our purpose of using the denotational semantics is to give precise
meaning to payment primitives, independent of its implementation (which deals with performance,
optimization, side effects, etc.). We will spend the part \ref{part:gpp} exploring the denotational
semantics for general payment primitives in modern payment system.

@@

The specifications will be formally defined using denotational semantics, along with a restatement
in Haskell also applied as a constructive artifact friendly to computer environment\footnote{One may
argue for a restatement in \textit{Agda} instead, for it has a richer dependently-type system for
desirable constructive proofs. This matter will be dealt in the ``future investigations'' section of
this paper.}.

@@

    \item Semantic function $[\![.]\!] :: \mathcal{M} \rightarrow [\![\mathcal{M}]\!] $ evaluates
the expression of money distribution, payment primitives, etc.
\end{itemize}

@@

To make it meaningful, it must satisfy the following law:

@@

\paragraph{Proportional Distribution Primitives}

@@

An even more general abstraction is to model participants involved a payment primitive
a \textit{network} (we use symbol $w$ for them).

@@

The answer is most likely a no. Not because it is not useful, after all mathematical formula is
about truth and perhaps also elegance in it, but because of we may miss optimization opportunities
needed in implementation when more specialized versions are used instead\footnote{We will not though
discuss the subjective aspects, \eg\ in software engineering principles such as YAGNI
(\url{https://en.wikipedia.org/wiki/You_aren't_gonna_need_it}), or user experience perspective of
the payment primitives.}.

In part \ref{part:sf}, we will look into a reference implementation where these optimizations are
made.

@@

\textit{Superfluid protocol} (``the protocol'')\footnote{The code base is available
at \url{https://github.com/superfluid-finance/protocol-monorepo/}.} is the first implementation of
the \textit{denotational semantics of payment primitives} (though before it was formalized and named
so) on \textit{Ethereum Virtual Machine} (\cite{wood2014ethereum}). The first version of the protocol
is written in Solidity programming language \footnote{\url{https://soliditylang.org/} - About
Solidity programming language.}.

To better serve as a reference implementation of the \textit{modern payment system} formalized by
this paper and its sequels, an implementation in Haskell was created. It aims to implement the full
specifications of (a) denotational semantics of payment primitives, (b) compositional financial
contracts for the payment primitives, (c) token \& note model of money mediums, and their execution
environment.

This paper covers the overview of the implementation with regards to (a).

@@

For the purpose of compositional financial contracts, it is useful to separate monetary values that
bearers can use immediately (called \textit{untappedValue} in code) from ones that are set aside for
other financial purpose. The concept of \textit{real time balance} is thus created. \textit{Real
time balance} is a functorful\footnote{Bartosz Milewski has an excellent series on
functors: \url{https://bartoszmilewski.com/2015/01/20/functors/}, where he uses the term
``functorful'' to convey the idea of generalized container.} of monetary values. It can be converted
to a single monetary value.

Since this is not the subject of this paper, it suffices to show its definition instead:

@@

\begin{itemize}
    \item preserve the program correctness (and if possible, with proof of equivalence),
    \item optimize for efficient computations,
    \item and provide a software interface for it.
\end{itemize}

@@

\subsection{Monetary Unit Data (MUD)}

Recall that all monetary units have a monetary value function: $ \nu :: u \rightarrow \mathbb{N}$.

@@

\begin{code}
class ( SuperfluidCoreTypes sft
      ) => MonetaryUnitDataClass mud sft | mud -> sft where
    -- | π function - balance provided (hear: π) by the monetary unit data.
    balanceProvided
        :: forall t rtb.
           -- indexed type aliases
           ( t ~ SFT_TS sft
           , rtb ~ SFT_RTB sft
           )
        => mud -> t -> rtb

-- | A semigroup constrained monetary unit data type class.
--
-- Note: a. ~mud~ that doesn't have a binary function may also be referred
--          to as "non-scalable" ~mud~.
--
--       a. What can make a ~mud~ "scalable" then is exactly when it is an
--          actual semigroup. Since a new state can be merged onto the
--          previous state to a new single state. It is still worth mentioning
--          that it is only a sufficient condition, since a monoid could still
--          "cheat" by linearly grow its data size on each binary operation.
class ( MonetaryUnitDataClass smud sft
      , Semigroup smud
      ) => SemigroupMonetaryUnitData smud sft
\end{code}

@@

\subsection{Agreement Contract}

@@

\begin{code}
-- | Agreement contract type class.
class ( SuperfluidCoreTypes sft
      , Default ac
      , MonetaryUnitDataClass ac sft
      , Traversable (AgreementOperationOutputF ac) -- <= Foldable Functor
      , Monoid (AgreementOperationOutput ac)
      ) => AgreementContract ac sft | ac -> sft where

    -- | ω function - apply agreement operation ~ao~ (hear: ω) to the agreement
    --                operation data ~ac~ to get a tuple of:
    --
    --   1. An updated ~ac'~.
    --
    --   2. A functorful delta of agreement monetary unit data ~mudsΔ~, which
    --      then can be appended to existing ~mudΔ~.  This is what can make an
    --      agreement scalable.
    applyAgreementOperation
        :: forall t ao aoo.
           -- indexed type aliases
           ( t ~ SFT_TS sft
           , ao ~ AgreementOperation ac
           , aoo ~ AgreementOperationOutput ac
           )
        => ac -> ao -> t -> (ac, aoo)

    -- | φ' function - functorize the existential semigroup monetary unit data
    --                 of agreement operation output
    functorizeAgreementOperationOutput
        :: forall any_smud muds f.
           ( any_smud `IsAnyTypeOf` MPTC_Flip SemigroupMonetaryUnitData sft
           , MonetaryUnitDataClass any_smud sft
           -- indexed type aliases
           , muds ~ AgreementOperationOutput ac
           , f ~ AgreementOperationOutputF ac
           )
        => Proxy any_smud
        -> muds -> f any_smud

    data family AgreementOperation ac :: Type
    data family AgreementOperationOutputF ac :: Type -> Type
    type family AgreementOperationOutput ac = (smuds :: Type) | smuds -> ac
\end{code}

@@

Figure \ref{fig:ac-omega} is a illustration of the $\omega$ function ``machinery''.

\begin{figure}[H]
    \centering
    \includegraphics[width=1\textwidth]{../assets/agreement-contract-omega-function.png}
    \caption{Agreement Contract $\omega$ Function}
    \label{fig:ac-omega}
\end{figure}

@@

It is also important to note that \textit{AgreementContract} itself is also
a \textit{MonetaryUnitDataClass}, but it is not a semigroup, hence it is used to replace the
previous agreement contract data. This has optimization implications, as in that if agreement
contracts do not produce zero balance then they must be included in the $\nu$ function, that can
make the $\nu$ function $O(N)$ to the number of agreement contracts a monetary unit is
associated with.

@@

\section{Lens Data Accessors}

@@

The pro-functor version of lens might seem very obscure at first, but its data representation is
rather self-explanatory: a lens is simply a pair of getter ($view$) and setter ($update$) encoded in
the data.

@@

\section{Useful Agreements}

\subsection{Instant Value MUD}

@@

Note that (a) lens \textbf{operator (\string^.)} is the view function (getter) of data, (b)
semigroup \textbf{operator ($<>$)} defines how two monetary unit data can be combined into one.

\subsection{Constant Flow Agreement (CFA) MUD}

@@

\subsection{Decaying Flow Agreement (CFA) MUD}

@@

\subsection{Universal Index (UIDX)}

@@

\subsection{Proportional Distribution Index (PDIDX)}

@@

\section{Hierarchy of Agreements}

To finish, figure \ref{fig:ac-agreement-data-structures} is the illustration of data structures for
the denotational payment primitives implemented with agreement framework.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{../assets/agreement-data-structures.png}
    \caption{Agreement Data Structures}
    \label{fig:ac-agreement-data-structures}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\part{Notes on Future Investigations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

One of the advantages of using Haskell language to write the reference implementation was its
industrial strength. Because of that, the implementation could be easily integrated into a
production code base without significant performance compromises.

However, one major disadvantage is that it is not equipped with sufficient apparatus for program
correctness proofs. The best it can offer without using more experimental Haskell features is to
use \textit{QuickCheck}\footnote{\url{https://wiki.haskell.org/QuickCheck} - Haskell wiki page for
QuickCheck.} to test the necessary laws are not violated using randomized test vector.

@@

It is based on the insight of the deep connection (equivalence/isomorphism) between logic and
dependently typed programming, often called "the Curry–Howard correspondence", as discovered and
developed during the 20th century (explained in \cite{wadler2015propositions} by Phillip Wadler as
``Propositions as Types''). Agda being a dependently typed functional programming language fully
embodies this insight, hence a good candidate for creating provable correct program.

To learn more about Agda, refer to these footnotes
\footnote{\url{https://plfa.github.io/} - Programming Language Foundations in Agda.}
\footnote{\url{https://wiki.portal.chalmers.se/agda/pmwiki.php} - Agda Wiki.}
\footnote{\url{https://wiki.portal.chalmers.se/agda/Main/Community} - Agda community page.}.

@@

The execution environments for these financial contracts are also fertile ground for new ideas, from
deterministic execution avoiding the usage of buffer, to using distributed ledger technology to
deep-embed these financial contracts in the consensus layer.

@@

The \textit{real time balance} introduced in this paper can be seen as a instance in
the \textit{general accounting domain} specifically for financial contracts in dealing with payment
primitives, which can be seen as a real time version of the cash flow accounting. Along with balance
sheet accounting, income statement accounting, the conversion between these instances of general
accounting domains are often called ``reconciliations''.

@@

\paragraph{Real Time Finance}

@@

This paper defines the what constitutes a modern payment system, with its payment primitives
formally specified using denotational semantics and restated using Haskell programming
language. Along with a reference implementation of the specification, we want to make a case that
marrying with the progress of computer science, the notation of money can perform its function of
medium of exchange can have a modern upgrade.

Absent of any claim on whether if such modern upgrade is needed, we ask the readers ask
ourselves a question together: with electronic money increasingly being used, should we keep
emulating the function money of its traditional quality, or could we rethink what it may be in the
information age?
